---
title: "Hate Crimes and Socioeconomic Factors in U.S. States"
subtitle: "A Statistical Analysis Using FBI and Census Data"
author: "Chingiz Rabdanov"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5, fig.align = "center")
```

# Abstract

This project investigates whether socioeconomic conditions in U.S. states are
associated with the rate of FBI-reported hate crimes per 100,000 residents.
Using a publicly available dataset compiled by FiveThirtyEight from the
American Community Survey (Census), the Kaiser Family Foundation, and the
FBI Uniform Crime Reporting program, we model the average annual hate-crime
rate in 2010–2015 as a function of two state-level predictors: median household
income and the Gini index of income inequality. A multiple linear regression
on the 50 states with complete data shows that both income (β = 7.42 × 10⁻⁵,
p = 0.002) and the Gini index (β = 39.36, p < 0.001) are positively and
significantly associated with the FBI hate-crime rate, jointly explaining
about 33% of the variation across states (R² = 0.33, F(2, 47) = 11.38,
p < 0.001). However, the relationship is highly sensitive to the District of
Columbia, an extreme outlier on both the dependent variable and inequality;
removing it reduces R² to 0.08 and renders the coefficients non-significant.
We therefore conclude that the headline correlation should be interpreted with
caution: state-level wealth and inequality are associated with reported hate
crimes, but reporting practices, population scale, and the influence of single
high-leverage jurisdictions complicate any causal interpretation. The analysis
illustrates how a single observation can dominate a small ecological dataset
and underscores the need for individual-level or longitudinal designs to
draw stronger inferences about the drivers of hate crime.

*(Word count: ~245)*

---

# Overview

## Context on the data collection

The dataset was assembled by FiveThirtyEight for the article *"Higher Rates of
Hate Crimes Are Tied to Income Inequality"* (Majumder, 2017). It combines:

- **American Community Survey (ACS, 2015)** — Census Bureau data on income,
  unemployment, education, and demographics.
- **Kaiser Family Foundation** — share of voters who voted for Donald Trump in
  the 2016 presidential election.
- **FBI Uniform Crime Reporting (UCR), 2010–2015** — average annual hate-crime
  rate per 100,000 residents.
- **Southern Poverty Law Center (SPLC), Nov 9 – Dec 18, 2016** — short-term
  count of hate incidents post-election.

Each row is a U.S. state (50 states + the District of Columbia, 51 rows total).

## Variables

**Dependent variable (DV) — quantitative**

- `avg_hatecrimes_per_100k_fbi`: Average annual hate-crime incidents per
  100,000 residents reported to the FBI between 2010 and 2015.

**Independent variables (IVs) — both quantitative**

1. `median_household_income`: Median household income in 2015 USD (ACS).
2. `gini_index`: 2015 Gini index of income inequality (0 = perfect equality,
   1 = one person earns everything).

We chose these two predictors because they capture two conceptually distinct
features of a state's economy: *level* of prosperity (income) and *distribution*
of prosperity (inequality). It is plausible that both shape inter-group tension
in different ways.

## Research question

> **To what extent do a state's median household income and its level of
> income inequality (Gini index) jointly predict the average annual rate of
> hate crimes per 100,000 residents reported to the FBI?**

## Type of study

This is an **observational, cross-sectional** ecological study. No variables
were manipulated; states are the units of analysis. Causal claims are not
warranted — we can only describe associations.

## Data source

Self-collected from a public mirror of FiveThirtyEight's
`hate_crimes.csv`:
`https://raw.githubusercontent.com/Chingiz1492/Hate-Crimes-US/refs/heads/main/hate_crimes.csv`

Original source article: Majumder, M. (2017). *Higher Rates of Hate Crimes
Are Tied to Income Inequality.* FiveThirtyEight.

---

# 1. Data preparation

```{r packages}
# Load packages (install once with install.packages() if needed)
library(tidyverse)   # dplyr, ggplot2, readr, etc.
library(broom)       # tidy() for model output
library(knitr)       # kable() for nice tables
library(GGally)      # ggpairs() for correlation matrix plot
```

```{r load-data}
url <- "https://raw.githubusercontent.com/Chingiz1492/Hate-Crimes-US/refs/heads/main/hate_crimes.csv"
hatecrime_raw <- read.csv(url)

# Quick look
glimpse(hatecrime_raw)
```

The raw data has **51 rows × 12 columns**. We keep only the variables relevant
to our research question and drop rows with missing values on those columns
(Hawaii has no FBI hate-crime figure).

```{r clean}
hate <- hatecrime_raw %>%
  select(state,
         hate_crimes = avg_hatecrimes_per_100k_fbi,
         income      = median_household_income,
         gini        = gini_index) %>%
  drop_na()

cat("Rows after dropping NA:", nrow(hate), "\n")
```

After cleaning we have **50 observations** (50 states + DC, minus Hawaii).

## Cases

Each case is one U.S. state (or DC). The unit of analysis is the *state*, not
the individual — this is what makes the design ecological.

---

# 2. Summary statistics

```{r summary-stats}
hate %>%
  select(hate_crimes, income, gini) %>%
  summary()
```

A more compact, presentation-ready table:

```{r summary-table}
hate %>%
  select(hate_crimes, income, gini) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(N      = n(),
            Mean   = mean(Value),
            SD     = sd(Value),
            Min    = min(Value),
            Median = median(Value),
            Max    = max(Value)) %>%
  kable(digits = 3, caption = "Summary statistics for the three core variables")
```

**Reading the table.** Across the 50 states, the average FBI hate-crime rate
is ≈ 2.37 per 100,000 (SD ≈ 1.71), but the spread is wide — from 0.27 (Wyoming)
to 10.95 (DC). Median income ranges from ≈ \$35.5K (Mississippi) to ≈ \$76K
(Maryland), and Gini ranges from 0.42 (Utah) to 0.53 (DC).

---

# 3. Visualizations

## 3.1 Distribution of the dependent variable

```{r dv-hist}
ggplot(hate, aes(x = hate_crimes)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  geom_vline(aes(xintercept = mean(hate_crimes)), linetype = "dashed") +
  labs(title = "Distribution of FBI hate-crime rate (per 100,000)",
       x = "Avg. hate crimes per 100k (2010–2015)",
       y = "Number of states") +
  theme_minimal()
```

The DV is right-skewed and has one striking observation on the far right
(DC, ≈ 11). This will matter later.

## 3.2 Scatter plot — income vs. hate crimes

```{r scatter-income}
ggplot(hate, aes(x = income, y = hate_crimes)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  geom_text(data = hate %>% filter(state == "District of Columbia"),
            aes(label = state), hjust = 1.1, color = "darkred") +
  labs(title = "Hate-crime rate vs. median household income",
       x = "Median household income (USD)",
       y = "FBI hate crimes per 100k") +
  theme_minimal()
```

There is a positive but modest trend; DC is clearly the upper-right outlier.

## 3.3 Scatter plot — Gini vs. hate crimes

```{r scatter-gini}
ggplot(hate, aes(x = gini, y = hate_crimes)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  geom_text(data = hate %>% filter(state == "District of Columbia"),
            aes(label = state), hjust = 1.1, color = "darkgreen") +
  labs(title = "Hate-crime rate vs. Gini index of inequality",
       x = "Gini index (higher = more unequal)",
       y = "FBI hate crimes per 100k") +
  theme_minimal()
```

Inequality shows a more visible positive association — but again DC drives it.

## 3.4 Correlation matrix

```{r corr-pairs, fig.height=5.5}
ggpairs(hate %>% select(hate_crimes, income, gini),
        title = "Pairwise relationships among the three variables")
```

---

# 4. Statistical analysis

## 4.1 Method

Both predictors and the outcome are quantitative, so we use **multiple linear
regression** (OLS). This lets us estimate the partial effect of each predictor
while controlling for the other.

## 4.2 Hypotheses

For each slope coefficient we test:

- **H₀:** βⱼ = 0  (the predictor has no linear association with the hate-crime
  rate after controlling for the other predictor)
- **Hₐ:** βⱼ ≠ 0  (there is a linear association)

For the overall model:

- **H₀:** β_income = β_gini = 0
- **Hₐ:** at least one slope is non-zero (overall F-test)

We use a significance level of α = 0.05 throughout.

## 4.3 Pearson correlations

```{r correlations}
cor_income <- cor.test(hate$income, hate$hate_crimes)
cor_gini   <- cor.test(hate$gini,   hate$hate_crimes)

tibble(Predictor = c("Median income", "Gini index"),
       r         = c(cor_income$estimate, cor_gini$estimate),
       `95% CI`  = c(sprintf("[%.3f, %.3f]",
                             cor_income$conf.int[1], cor_income$conf.int[2]),
                     sprintf("[%.3f, %.3f]",
                             cor_gini$conf.int[1],   cor_gini$conf.int[2])),
       `p-value` = c(cor_income$p.value, cor_gini$p.value)) %>%
  kable(digits = 4, caption = "Pearson correlations with FBI hate-crime rate")
```

## 4.4 Multiple regression model

```{r model}
model <- lm(hate_crimes ~ income + gini, data = hate)
summary(model)
```

A tidy version for reporting:

```{r model-tidy}
tidy(model, conf.int = TRUE) %>%
  kable(digits = 5,
        caption = "Coefficient estimates with 95% confidence intervals")

glance(model) %>%
  select(r.squared, adj.r.squared, statistic, p.value, df, df.residual, nobs) %>%
  kable(digits = 4, caption = "Model fit statistics")
```

### Interpretation

- **Intercept (β₀ ≈ −19.59).** Not directly meaningful — it is the predicted
  hate-crime rate when *both* income and Gini are zero, which is outside the
  data range.
- **Income (β ≈ 7.42 × 10⁻⁵, p = 0.002).** Holding inequality constant, every
  additional \$10,000 of median household income is associated with about
  **+0.74** additional hate crimes per 100,000 residents. Counter-intuitive
  on its own — wealthier states report *more* incidents — but plausibly
  reflects better reporting infrastructure rather than more bias.
- **Gini index (β ≈ 39.36, p < 0.001).** Holding income constant, a 0.01
  increase in the Gini index is associated with about **+0.39** additional
  hate crimes per 100,000.
- **Overall F-test:** F(2, 47) = 11.38, p ≈ 9 × 10⁻⁵ — we reject the joint
  null; the model explains a non-trivial share of variation.
- **R² = 0.326** — the two predictors together account for about **32.6%** of
  the variance in the FBI hate-crime rate across states.
  **Adjusted R² = 0.298**, which corrects for the two predictors and gives a
  fairer estimate (about 30%). Roughly two-thirds of the variation between
  states is *not* explained by income and inequality alone.

## 4.5 Diagnostics

```{r diagnostics, fig.height=6, fig.width=8}
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

The residuals-vs-fitted and Q-Q plots reveal one observation with both very
high leverage and a very large residual: **the District of Columbia**. The
Cook's distance plot confirms it sits far above the standard threshold.

## 4.6 Sensitivity analysis — refit without DC

```{r model-no-dc}
model_no_dc <- lm(hate_crimes ~ income + gini,
                  data = hate %>% filter(state != "District of Columbia"))

tidy(model_no_dc, conf.int = TRUE) %>%
  kable(digits = 5,
        caption = "Regression without the District of Columbia")

glance(model_no_dc) %>%
  select(r.squared, adj.r.squared, statistic, p.value, nobs) %>%
  kable(digits = 4, caption = "Fit statistics (DC removed)")
```

**This is the most important diagnostic in the project.** Without DC:

- R² drops from **0.33 → 0.08**.
- The income coefficient is no longer significant (p ≈ 0.06).
- The Gini coefficient is no longer significant (p ≈ 0.32).

DC is unique: it is a single city treated as a "state", so it has both an
extreme Gini and an extreme hate-crime rate. The headline correlation is
largely driven by this single observation.

---

# 5. Conclusion

## What we found

Across the 50 U.S. states with complete data, both median household income
and the Gini index are positively associated with FBI-reported hate-crime
rates, and together they explain about a third of the variation between
states (R² ≈ 0.33). The overall model is statistically significant
(F(2, 47) = 11.38, p < 0.001), and each predictor remains significant in the
presence of the other — suggesting that *level* of income and *distribution*
of income capture different aspects of state-level conditions.

## Why this matters

Hate crimes have severe psychological, social, and political consequences. If
states with greater economic inequality report systematically more hate
crimes, that is relevant evidence for policymakers, civil-rights
organizations, and journalists, and it aligns with social-strain theories
in sociology. Even the *reporting* angle is policy-relevant: if richer states
just record more incidents, then poorer states' figures may understate the
true problem.

## Limitations

1. **Outlier dominance.** The District of Columbia drives most of the
   relationship. With DC removed, neither predictor is significant. Findings
   should be interpreted as descriptive of *this particular sample including
   DC*, not as a robust population-level effect.
2. **Ecological fallacy.** State-level associations do not necessarily hold at
   the individual level. We cannot claim that wealthier or more unequal
   *people* commit more hate crimes.
3. **Reporting bias.** FBI hate-crime statistics depend on whether local law
   enforcement participates in the UCR program and how they classify incidents.
   Apparent state-to-state differences may partly reflect reporting practices
   rather than actual incidence.
4. **Cross-sectional design.** With one cross-section we cannot assess whether
   changes in inequality lead to changes in hate crime; reverse causality and
   confounders (urbanization, demographic composition, political climate) are
   unaddressed here.
5. **Small N (50).** State-level regressions have limited statistical power
   and are sensitive to single observations, as we saw with DC.

## Suggested next steps

- Use longitudinal panel data (state × year) and fixed effects.
- Add control variables (urbanization, racial composition, share of college
  graduates, hate-crime law strength).
- Replicate with a different DV (e.g., SPLC counts, ADL data) for triangulation.
- Consider robust regression or a log-transformed DV to mitigate the influence
  of DC and other high-leverage points.

---

# Session info

```{r session}
sessionInfo()
```
